\bf Promoting analysis reproducibility with accessibility: An example in evolutionary biology

### 2022-02-03 jsdse/Manuscript.tex
adding to conclusions

\newpage
\spacingset{1.45} % DON'T change the spacing!
\section{Introduction}
\label{sec:intro}

@@

\noindent\fbox{%
    \parbox{\textwidth}{%
    \textbf{Box 1.}
      Availability does not imply accessibility. One example we find really useful is the
      marshmallows in an office story.
      Brand new marshmallow bags remained in a common office area after a social gathering.
      The marshmallows did not belong to anyone anymore, so they were freely
      available for eating by anyone in the office.
      Yet, the marshmallow bags stayed in the common area, unopened for days and even weeks.
      It was not until someone opened the bags and placed the marshmallows on a tray,
      that people started actually eating them. They were gone in a matter of hours.
      Scientific products are not sweet as marshmallows, but they are a coveted asset.
      While sharing, describing and documenting makes scientific products available,
      it does not necessarily makes them accessible. For example, documentation of an
      available scientific workflow might be written using expert language that is too foreign
      for a general audience to successfully reproduce the workflow. While some individuals might
      be able to invest the time to learn the expert language to successfully reproduce
      the workflow, this will not be the case for the majority, even if it is
      something that would be useful long term, the short term time investment
      is often too high for individuals to engage with it.
    }%
}
\bigskip

To elaborate on our thesis, we use an example in evolutionary biology that focuses
on understanding the shared ancestry among organisms through phylogenetic trees, which provide the
basis to study and understand biological processes in an evolutionary context (\cite{dobzhansky1973nothing}).
As such, improving reproducibility and availability
of phylogenetic research is relevant for many aspects of biological research.
The Open Tree of Life project (OpenTree) has developed a platform that facilitates
availability of results from phylogenetic research, by standardizing and storing
phylogenetic data with the goal of synthesizing a single phylogenetic tree encompassing
all life (\cite{opentreeoflife2019synth}).
All data in OpenTree is open access and available programmatically through its many
Application Programming Interface (API) services (\cite{opentreeAPIs}).
Additionally, R packages (\cite{michonneau2016rotl}) and Python libraries
(\cite{mctavish2021opentree}) have been developed as wrappers for OpenTree API services
to make them available to a wider programming audience.
The R and Python OpenTree wrappers have been utilized by computer-literate individuals,
to seamlessly establish reproducible workflows to use and reuse expert phylogenetic
knowledge for biological research (\cite{sanchez2019datelife}) and education
(\cite{nguyen2020phylotastic, phylotasticedtools, galacticedtools}).

@@

\section{Methods}
\label{sec:meth}
\subsection*{Identifying hurdles to accessibility}

@@

\subsection*{Adressing hurdles to accessibility: the principles}

\subsubsection*{1. Demonstrate integration of function usage with motivating examples}

@@

\subsubsection*{2. Demonstrate errors and warnings thoroughly}

@@

We identify ways to evaluate inputs to know if they will produce an error, and design
alternatives on what to do when faced with an error or warning, and demonstrate
these alternatives.
One of the most essential skills in programming is interpreting and moving forward
from errors.
Many finely honed tutorials do not trigger errors, which precludes helping students
to develop the tools to understand and address errors when they do encounter them,
as they inevitably will.

@@

\subsubsection*{3. Avoid jargon and expert language}

@@

\subsubsection*{4. Make it stable through time}

@@

\section{Conclusion}
\label{sec:conclusion}

We explain the warnings and errors and design ways to avoid them, and detect them
 beforehand (i.e., before using an input that would give an error).
  We explain the consequences of warnings.
We designed ways to access the different elements of the outputs.
We have received emails from senior researchers thanking us for this materials,
 and students have been able to engage using the packages with less hep from the PIs.

@@

Intentional integration of the subject of scientific reproducibility into
 the curriculum is a key aspect to promote reproducible practices long term.
There are different elements that permit fully reproducible science.

>>>>

\newpage
\spacingset{1.45} % DON'T change the spacing!
\section*{Introduction}
\label{sec:intro}

@@

\noindent\fbox{%
    \parbox{\textwidth}{%
    \textbf{Box 1. The role of accessibility to achieve full availability.}
      Availability is generally defined as the ability or potential of something
      to be used and accessed (\cite{availability2022cambridge}).
      Following this definition, accessibility is generally considered a synonym of availability.
      Yet, in practice, availability does not imply accessibility.
      One example we find really useful to practically differentiate the concepts of availability and accessibility
      is the "marshmallows in an office" story, which goes like this.
      A couple of brand new marshmallow bags remained in a common office area after
      a social gathering.
      The marshmallows were not claimed by anybody, so they were left in the common are,
      freely available to be taken or eaten by anyone in the office.
      Yet, the marshmallow bags stayed in the common area, unopened for days and even weeks.
      It was not until someone decided to open the bags and placed the marshmallows on a tray,
      that people started actually eating them. They were gone in a matter of hours.
      Scientific products are not sweet as marshmallows, but they are a coveted resource.
      While sharing, describing and documenting makes scientific products available,
      it does not necessarily makes them accessible. For example, documentation of an
      available scientific workflow might be written using expert language that is too foreign
      for a general audience to successfully reproduce the workflow. While some individuals might
      be able to invest the time to learn the expert language to successfully reproduce
      the workflow, this will not be the case for the majority, even if it is
      something that would be useful long term, the short term time investment
      is often too high for individuals to engage with it, making the workflow and its results
      effectively inaccessible and hence unavailable to the majority of the community.
    }%
}
\bigskip

To elaborate on our thesis, we use an example in evolutionary biology that focuses
on understanding the shared ancestry of organisms through phylogenetic trees, which provide the
basis to study and understand biological processes in an evolutionary context (\cite{dobzhansky1973nothing}).
As such, improving reproducibility and availability
in phylogenetic research is relevant for many aspects of biological research.
The Open Tree of Life project (OpenTree) has developed a platform that facilitates
availability of results from phylogenetic research, by standardizing and storing
phylogenetic data with the goal of synthesizing a single phylogenetic tree encompassing
all life (\cite{opentreeoflife2019synth}).
All data in OpenTree is open access and available programmatically through its many
Application Programming Interface (API) services (\cite{opentreeAPIs}).
Additionally, R packages (\cite{michonneau2016rotl}) and Python libraries
(\cite{mctavish2021opentree}) have been developed as wrappers for OpenTree API services
to make them available to a wider programming audience.
The R and Python OpenTree wrappers have been utilized by computer-literate individuals,
to seamlessly establish reproducible workflows to use and reuse expert phylogenetic
knowledge for biological research (\cite{sanchez2019datelife}) and education
(\cite{nguyen2020phylotastic, phylotasticedtools, galacticedtools}).

@@

\section*{Identifying hurdles to accessibility}
\label{sec:identifying}

@@

\section*{Addressing hurdles to accessibility: some principles}
\label{sec:addressing}

\subsubsection*{a. Demonstrate integration of function usage with motivating examples}

@@

\subsubsection*{b. Demonstrate errors and warnings thoroughly}

@@

We identify ways to evaluate inputs to know if they will produce an error, and design
alternatives on what to do when faced with an error or warning, and demonstrate
these alternatives.
One of the most essential skills in programming is interpreting and moving forward
from errors.
Many finely honed tutorials do not trigger errors, which precludes helping students
to develop the tools to understand and address errors when they do encounter them,
as they inevitably will.
On our tutorial, we focus on explaining the meaning and downstream coof warnings and errors, and
 showcase ways to detect them before they are triggered (i.e., before using an input
  that would elicit a warning or error). This has two pedagogical benefits:
1) it provides users/students with the means to troubleshoot their own warnings and erros, and
2) it allows users/students to understand with more depth what the function is doing.

We designed ways to access the different elements of the outputs.

@@

\subsubsection*{c. Avoid jargon and expert language}

@@

\subsubsection*{d. Make it stable through time}

@@

\section*{Conclusion}
\label{sec:conclusion}

Ultimately, the improvement of reproducibility rates in science long term depends
on our ability to intentionally integrate the subject of reproducibility in the
the curriculum for college students and aspiring researchers to develop the skills
necessary to create reproducible scientific workflows and materials.

Some universities have been incorporating the subject in their classes (see University
 of Washington and The National Institute for General Medical Sciences).
The focus has been on students to develop skills to document their work.
The principles identified and outlined here can be used to set learning goals and
 outcomes on new reproducibility syllabus.

We have received emails from senior researchers thanking us for this materials,
 and students have been able to engage using the packages with less hep from the PIs.

@@
